Finally Tagless, Partially Evaluated

### 2008-09-17 tagless/tagless-final.tex
a fair number of changes, but mostly dealing with bibliographic issues as well as making our contributions clearer.  Removed section 6 in its entirety but added no comments about the topic.

%% TODO: add a ref 
%%  Lars Birkedal's master thesis
%%   http://www.itu.dk/people/birkedal/papers/paresm.ps.gz

@@

\begin{abstract}
\ifdim\parindent=0pt \parindent=1em \fi
We have built the first family of tagless interpretations for a
higher-order typed object
language in a typed metalanguage (Haskell or ML) that
require no dependent
types, generalized algebraic data types, or
postprocessing to eliminate tags.
The statically type-preserving interpretations include
an evaluator, a compiler (or staged evaluator), a partial
evaluator, and call-by-name and
call-by-value CPS transformers.

Our main idea is to encode
\ifshort HOAS \else de Bruijn or higher-order abstract syntax \fi
using combinator functions rather than data constructors.
In other words, we represent object terms not in an initial algebra
but using the coalgebraic structure of the $\lambda$-calculus.
Our representation also simulates inductive maps from types to
types, which are required for typed partial evaluation and CPS transformations.

Our encoding of an object term abstracts over the various ways to
interpret it, yet statically assures that the interpreters never get
stuck.  To achieve self\hyp interpretation and show Jones\hyp
optimality, we relate this exemplar of higher-rank and higher-kind
polymorphism \ifshort\else (provided by ML functors and Haskell~98 constructor
classes) \fi to plugging a term into a context of let\hyp polymorphic
bindings.
\end{abstract}

@@

We emphasize ``typed'' and ``fold'' in the previous sentence.
Our encoding of the recursive type of terms \citep{bohm-automatic}
makes folds over terms much easier to write than functions over
terms that are not primitive recursive (or, compositional).
In contrast, \citearound{'s encoding of the sum type of
terms}\citet{JFP-Mogensen} does not privilege folds.
In exchange, we statically express object types in the metalanguage
and prevent both kinds of run-time errors in
\S\ref{tagproblem}, due to evaluating ill-typed or open terms.
Because the new interpreter
uses no universal type or pattern matching, it never gives a
run-time error, and is in fact total.  Because this safety is obvious
not just to us but also to the metalanguage implementation, we avoid
the serious performance penalty \citep{WalidICFP02}
that inevitably arises \citep{Gluck-jones-optimality}
of error checking at run time.

@@

We attack the problem of tagless (staged) type-preserving
interpretation exactly as it was posed
by \citet{WalidICFP02} and \citet{xi-guarded}.
We use their running examples and achieve the
result they call desirable.  Our contributions are as follows.
\begin{enumerate}
\item We build the first \emph{family} of interpreters, each instantiating the
   \emph{same} signature, that evaluate (\S\ref{language}),
   compile (\S\ref{S:compiler}), and 
   partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing
   styles\ifshort\else\ (\S\ref{variations})\fi.
\item These interpreters use no type tags
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
    Yet the type system of the metalanguage
    assures statically that each object program is well-typed and closed,
    and that each interpreter preserves types and never gets stuck.
    In particular, our (online) partial evaluator and CPS transformers
    avoid GADTs in their implementation and stay portable across
    Haskell 98 and ML, by expressing in their interface an inductive
    map from input types to output types.
\item Our clean, comparable implementations using OCaml modules and Haskell
    type classes show how to parametrize our final representation of object
    terms over multiple ways to assign them meanings.
\item We point a clear way to extend the object language with more features
    such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
    \ Our term encoding is contravariant in the object language, so
    extending the language does not invalidate terms already encoded.
\item We describe an approach to self\hyp interpretation compatible with the
  above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
  out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
\end{enumerate}
Our code is surprisingly simple and obvious in hindsight, but
it has been cited as a difficult problem (\cite{sumii-hybrid}
notwithstanding) to interpret a typed object language in a typed metalanguage
without tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

For brevity, we just
use the module returned by |CPST| as is. Because it does not
match the signature |Symantics|, we cannot apply the |EX| functor to it.
Nevertheless, we can write the tests.
\begin{code}
module T = struct
  module M = CPST(C)
  open M
  let test1 () =
       app (lam (fun x -> x)) (bool true)  (* same as before *)
  let testpowfix () = ...                  (* same as before *)
  let testpowfix7 =                        (* same as before *)
       lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code}
We instantiate |CPST| with the desired base interpreter~|C|,
then use the result |M| to
interpret object terms. Those terms are \emph{exactly} as before.
Having to textually copy the terms is the
price we pay for this simplified treatment.
Our discussion of self\hyp interpretation in~\S\ref{selfinterp} shows
that this copying is not frivolous but represents plugging a term into
a context, which is one of the many faces of polymorphism.

@@

\ifshort\else
\section{Self-interpretation}\label{selfinterp}

@@

In the presence of let-bound polymorphism, we can understand a term
waiting to be plugged into a context as a higher-rank and higher-kind
abstraction over the context.  Even though our object language does not support
higher abstraction, our metalanguages do, so they can type-check an object term
separately from its interpreter---either as a functor from a |Symantics| module
containing a type constructor
(in OCaml), or a value with a |Symantics| constraint over a type
constructor (in Haskell).  Thus,
``context'' is a euphemism for a polymorphic argument, and ``plugging''
is a euphemism for application.

@@

@@

\ifshort
Our initial motivation came from several papers 
\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple}
that use embedded interpreters to justify advanced
type systems, in particular GADTs.
\else
Our initial motivation came from several papers that justify advanced
type systems, in particular GADTs, by embedded interpreters
\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} and
CPS transformations \citep{Guillemette-Monier-PLPV,shao-type-toplas,chen-typeful}.
\fi
We admire all this technical machinery, but
these motivating examples do not need it.
Although GADTs may indeed be simpler and more flexible, they are
unavailable in mainstream ML, and their implementation in GHC
6.6.1 fails to
detect exhaustive pattern matching.  We also wanted to find the minimal
set of widespread language features needed for tagless
type-preserving interpretation.

@@

@@

Our separation between the |Symantics| interface and its many
implementations codifies the common practice of implementing an embedded
DSL by specifying an abstract syntax of object\hyp language pervasives,
such as addition and application, then providing multiple
interpretations of them.  \Citet{JonesNielsen-D58} prefigured this
separation when they decomposed a denotational definition of an untyped
object language into a core semantics (which we call abstract syntax)
and multiple interpretations.  When implementing an embedded DSL in this
way, it is also common practice to use phantom types to rule out
ill-typed object terms, as done in Lava
\citep{Lava} and by \citet{Rhiger-thesis}. However, these two approaches
are not tagless because they still use universal types, such as Lava's
\texttt{Bit} and \texttt{NumSig}, and Rhiger's \texttt{Raw} (his Figure~2.2)
and \texttt{Term} (his Chap.~3), which incur the attendant overhead of
pattern matching.  The universal type also
greatly complicates the soundness and completeness proofs of embedding
\citep{Rhiger-thesis}, whereas our proofs are trivial.
Rhiger's approach does not support typed CPS transformation (his~\S3.3.4).
\begin{comment}
Rhiger's But Fig 2.2, p33: universal type Raw.  He uses phantom type
upon the Exp datatype. But that is cheating: phantom type means
essentially we can easily do coerce. We use real types.  That's why he
had to do tedious proofs in Sec 2 of soundness and completeness of
embedding. Whereas our proofs are obvious.  His sec 3 is based on data
representation of terms. They have type tags.  We do nothing of that
kind: See Sec 3.1.2. See numerous "data Term" in Sec3, which is the U
type.  In Sec 3.3.4 (p76) Rhiger specifically says that his encoding
cannot do typed CPS transformation -- whereas our does. BTW, Rhiger
thesis contains the definitions of the interpreter and the compiler,
in the beginning. Use this in response to Rev1)
\end{comment}

We are not the first to implement a typed interpreter for a typed
language.  \Citet{laod93} use type classes to implement a metacircular
interpreter (rather than a self\hyp interpreter) of a
typed version of the SK language, which is quite different from our
object language.  Their interpreter
appears to be tagless, but they could not have implemented a
compiler or partial evaluator in the same way, since they rely
heavily on injection\hyp projection pairs.

@@

We could not find work that establishes that
the \emph{typed} $\lambda$-calculus has a final coalgebra structure.
\ifshort
(See \Citet{honsell99coinductive} for the untyped case.)
\else
\Citet{HonsellLenisa,honsell99coinductive}
investigate the untyped $\lambda$-calculus
along this line.  
In particular, they use
contexts with a hole \citep[p.\,13]{honsell99coinductive} to define
\emph{observational equivalence}
(see our~\S\ref{selfinterp}).
\citearound{'s bibliography}\Citet{honsell99coinductive} refers to the
foundational work in this important area.  
Particularly intriguing is the link to the
coinductive aspects of B\"{o}hm trees, as pointed out by
\citet{berarducci-models} and Jacobs \citeyearpar[Example 4.3.4]{jacobs-coalgebra}.
\fi

\ifshort We observe that \else As \S\ref{S:clarify} observes, \fi
higher-rank and higher-kind
polymorphism lets us type-check and compile object terms separately from
interpreters.  This \ifshort\else observation \fi is consistent with the role of
polymorphism in the separate compilation of modules
\citep{shao-typed}.

>>>>

@@

\begin{abstract}
\ifdim\parindent=0pt \parindent=1em \fi
We have built the first \emph{family} of tagless interpretations for a
higher-order typed object
language in a typed metalanguage (Haskell or ML) that
require no dependent
types, generalized algebraic data types, or
postprocessing to eliminate tags.
The statically type-preserving interpretations include
an evaluator, a compiler (or staged evaluator), a partial
evaluator, and call-by-name and
call-by-value CPS transformers.

Our principal technique is to encode
\ifshort HOAS \else de Bruijn or higher-order abstract syntax \fi
using combinator functions rather than data constructors.
In other words, we represent object terms not in an initial algebra
but using the coalgebraic structure of the $\lambda$-calculus.
Our representation also simulates inductive maps from types to
types, which are required for typed partial evaluation and CPS transformations.

Our encoding of an object term abstracts over the various ways to
interpret it, yet statically assures that the interpreters never get
stuck.  
%% To achieve self\hyp interpretation and show Jones\hyp
%% optimality, we relate this exemplar of higher-rank and higher-kind
%% polymorphism \ifshort\else (provided by ML functors and Haskell~98 
%% constructor classes) \fi to plugging a term into a context of 
%% let\hyp polymorphic bindings.

As a side-effect, we provide another convincing application of the 
usefulness of being able to abstract over higher-kinded types.
\end{abstract}

@@

We emphasize ``typed'' and ``fold'' in the previous sentence.
Our encoding of the recursive type of terms \citep{bohm-automatic}
makes folds over terms much easier to write than functions over
terms that are not primitive recursive (or, compositional).
In contrast, \citearound{'s encoding of the sum type of
terms}\citet{JFP-Mogensen} does not privilege folds.
In exchange, we statically express object types in the metalanguage
and prevent both kinds of run-time errors in
\S\ref{tagproblem}, due to evaluating ill-typed or open terms.
Because the new interpreter
uses no universal type or pattern matching, it never gives a
run-time error, and is in fact total.  Because this safety is obvious
not just to us but also to the metalanguage implementation, we avoid
the serious performance penalty \citep{WalidICFP02}
that arises from error checking at run time.

@@

We attack the problem of tagless (staged) type-preserving
interpretation exactly as it was posed
by \citet{WalidICFP02} and \citet{xi-guarded}.
We use their running examples and achieve the
result they call desirable.  Our contributions are as follows.
\begin{enumerate}
\item We build the first \emph{family} of interpreters, each instantiating the
   \emph{same} signature, that evaluate (\S\ref{language}),
   compile (\S\ref{S:compiler}), and 
   partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing
   styles\ifshort\else\ (\S\ref{variations})\fi.
\item These interpreters use no type tags
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
    Yet the type system of the metalanguage
    assures statically that each object program is well-typed and closed,
    and that each interpreter preserves types and never gets stuck.
    In particular, our (online) partial evaluator and CPS transformers
    avoid GADTs in their implementation and stay portable across
    Haskell 98 and ML, by expressing in their interface an inductive
    map from input types to output types.
\item Our clean, comparable implementations using OCaml modules and Haskell
    type classes show how to parametrize our final representation of object
    terms over multiple ways to assign them meanings.
\item We point a clear way to extend the object language with more features
    such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
    \ Our term encoding is contravariant in the object language, so
    extending the language does not invalidate terms already encoded.
%% \item We describe an approach to self\hyp interpretation compatible with the
%%   above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
%%   out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
\item Show how higher-kinded abstraction can be useful for building
    embedded DSLs.
\end{enumerate}
Our code is surprisingly simple and obvious in hindsight, but
it has been cited as a difficult problem (\cite{sumii-hybrid}
and \cite{Thiemann-combinators}
notwithstanding) to interpret a typed object language in a typed metalanguage
without tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

For brevity, we just
use the module returned by |CPST| as is. Because it does not
match the signature |Symantics|, we cannot apply the |EX| functor to it.
Nevertheless, we can write the tests.
\begin{code}
module T = struct
  module M = CPST(C)
  open M
  let test1 () =
       app (lam (fun x -> x)) (bool true)  (* same as before *)
  let testpowfix () = ...                  (* same as before *)
  let testpowfix7 =                        (* same as before *)
       lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code}
We instantiate |CPST| with the desired base interpreter~|C|,
then use the result |M| to
interpret object terms. Those terms are \emph{exactly} as before.
Having to textually copy the terms is the
price we pay for this simplified treatment.
\begin{comment}
Our discussion of self\hyp interpretation in~\S\ref{selfinterp} shows
that this copying is not frivolous but represents plugging a term into
a context, which is one of the many faces of polymorphism.
\end{comment}

@@

\ifshort\else
\begin{comment}
\jacques{Should we really eliminate this entirely?}
\section{Self-interpretation}\label{selfinterp}

@@

In the presence of let-bound polymorphism, we can understand a term
waiting to be plugged into a context as a higher-rank and higher-kind
abstraction over the context.  Even though our object language does not support
higher abstraction, our metalanguages do, so they can type-check an object term
separately from its interpreter---either as a functor from a |Symantics| module
containing a type constructor
(in OCaml), or a value with a |Symantics| constraint over a type
constructor (in Haskell).  Thus,
``context'' is a euphemism for a polymorphic argument, and ``plugging''
is a euphemism for application.
\end{comment}

@@

\jacques{From what other people have written (like those who refer to our
work), it seems these comments of ours were prescient and not altogether
obvious.  We should consider adding some comments about the crucial role
of 'repr' (r below) somewhere in the paper}

@@

\ifshort
Our initial motivation came from several papers 
\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple}
that use embedded interpreters to justify advanced
type systems, in particular GADTs.
\else
Our initial motivation came from several papers that justify advanced
type systems, in particular GADTs, by embedded interpreters
\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} and
CPS transformations \citep{Guillemette-Monier-PLPV,shao-type-toplas,chen-typeful}.
\fi
We admire all this technical machinery, but
these motivating examples do not need it.
Although GADTs may indeed be simpler and more flexible, they are
unavailable in mainstream ML, and their implementation in GHC
6.8 fails to
detect exhaustive pattern matching.  We also wanted to find the minimal
set of widespread language features needed for tagless
type-preserving interpretation.

@@

That it is possible to regard our various evaluators as forming a 
family is also an old idea \citep{Holst-AMIX}. \Citeauthor{Holst-AMIX}
introduces \emph{language triplets} as an interesting mechanism to 
relate (untyped) languages in much the same way we do.  Seeds of 
several of the techniques we use here can also be found in this paper.

@@

Our separation between the |Symantics| interface and its many
implementations codifies the common practice of implementing an embedded
DSL by specifying an abstract syntax of object\hyp language pervasives,
such as addition and application, then providing multiple
interpretations of them.  \Citet{JonesNielson:94:AbstractInterpretation}
prefigured this separation when they decomposed a denotational definition of an
untyped object language into a core semantics (which we call abstract syntax)
and multiple interpretations.  When implementing an embedded DSL in this
way, it is also common practice to use phantom types to rule out
ill-typed object terms, as done in Lava
\citep{Lava} and by \citet{Rhiger-thesis}. However, these two approaches
are not tagless because they still use universal types, such as Lava's
\texttt{Bit} and \texttt{NumSig}, and Rhiger's \texttt{Raw} (his Figure~2.2)
and \texttt{Term} (his Chap.~3), which incur the attendant overhead of
pattern matching.  The universal type also
greatly complicates the soundness and completeness proofs of embedding
\citep{Rhiger-thesis}, whereas our proofs are trivial.
Rhiger's approach does not support typed CPS transformation (his~\S3.3.4).
\begin{comment}
Rhiger's But Fig 2.2, p33: universal type Raw.  He uses phantom type
upon the Exp datatype. But that is cheating: phantom type means
essentially we can easily do coerce. We use real types.  That's why he
had to do tedious proofs in Sec 2 of soundness and completeness of
embedding. Whereas our proofs are obvious.  His sec 3 is based on data
representation of terms. They have type tags.  We do nothing of that
kind: See Sec 3.1.2. See numerous "data Term" in Sec3, which is the U
type.  In Sec 3.3.4 (p76) Rhiger specifically says that his encoding
cannot do typed CPS transformation -- whereas our does. BTW, Rhiger
thesis contains the definitions of the interpreter and the compiler,
in the beginning. Use this in response to Rev1
\end{comment}

In many ways, our work builds on that of \citep{Thiemann-combinators}.
We use the same basic techniques, but our partial evaluator is online
as this is the only way to make partial evaluation ``fit'' in our
family.  In other words, by moving to an online partial evaluator and
using staging instead of a $2$-level language, we are able to show that
\citeauthor{Thiemann-combinators}'s techniques are highly parametric
and typeable using mainstream features.  Using many different type
classes, including constructor classes, \citep{Thiemann-GenClass}
presents a Gofer implementation of set of \emph{binding-time polymorphic}
combinators.  By gathering all their classes into one and dropping the
requirement for a polymorphic |lift|, they could have invented
|Symantics| themselves.

We are not the first to implement a typed interpreter for a typed
language.  \Citet{laod93} use type classes to implement a metacircular
interpreter (a special kind of self\hyp interpreter) of a
typed version of the SK language, which is quite different from our
object language.  Their interpreter
appears to be tagless, but they could not have implemented a
compiler or partial evaluator in the same way, since they rely
heavily on injection\hyp projection pairs.

@@

We could not find work that establishes that
the \emph{typed} $\lambda$-calculus has a final coalgebra structure.
\ifshort
(See \Citet{honsell99coinductive} for the untyped case.)
\else
\Citet{HonsellLenisa,honsell99coinductive}
investigate the untyped $\lambda$-calculus
along this line.  
\begin{comment}
In particular, they use
contexts with a hole \citep[p.\,13]{honsell99coinductive} to define
\emph{observational equivalence}
(see our~\S\ref{selfinterp}).
\end{comment}
\citearound{'s bibliography}\Citet{honsell99coinductive} refers to the
foundational work in this important area.  
Particularly intriguing is the link to the
coinductive aspects of B\"{o}hm trees, as pointed out by
\citet{berarducci-models} and Jacobs \citeyearpar[Example 4.3.4]{jacobs-coalgebra}.
\fi

We observe that
higher-rank and higher-kind
polymorphism lets us type-check and compile object terms separately from
interpreters.  This \ifshort\else observation \fi is consistent with the role of
polymorphism in the separate compilation of modules
\citep{shao-typed}.
